---
title: "Interactive tables in Quarto"
format: html
---

You can create complex data visualization with R and Quarto, but there may be times when you just want to show your table of data in a way that is useful to readers. There are a number of ways to format tables. This page provides a preview of what you can do with tables in Quarto.

If you want a nicely formatted static table, look at the [gt package](https://gt.rstudio.com). The documentation for the package is very thorough and provides many examples.

A minimally interactive option is `rmarkdown::paged_table()`, which can be chosen as the default table output within the [`df-print` option in Quarto's YAML](https://quarto.org/docs/computations/r.html#data-frames). But you can get even fancier with either the [DT package](https://rstudio.github.io/DT/) or [reactable](https://glin.github.io/reactable/).

## Examples
Let's show examples using the movements of two sibling groups at the end of the 16th century.

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(DT)
library(reactable)
```

```{r}
#| label: load-data
#| message: false
#| echo: false
movements <- read_csv("../datasets/movements.csv")
locations <- read_csv("../datasets/locations.csv")
```

```{r}
#| label: load-data-show
#| eval: false
movements <- read_csv("https://raw.githubusercontent.com/jessesadler/vt5444s26/refs/heads/main/datasets/movements.csv")
```

### Regular tibble output

```{r}
#| label: tibble-output
movements
```

### RMarkdown paged

```{r}
#| label: paged-output
rmarkdown::paged_table(movements)
```

### DT

```{r}
#| label: dt-output
datatable(movements)
```

### reactable

```{r}
#| label: reactable-output
reactable(movements)
```

## reactable
The [DT package](https://rstudio.github.io/DT/) provides a nice default with a search box and includes many [options](https://rstudio.github.io/DT/options.html) for styling the table. However, [reactable](https://glin.github.io/reactable/) takes this even further, and so that is the package we will concentrate on here.

### Resources and examples
The below will show some examples to help get you started, but look at the [documentation](https://glin.github.io/reactable/) for the many options available to you.

- Scroll through the [Examples](https://glin.github.io/reactable/articles/examples.html) to get an idea of the basics.
- Go through the [demos](https://glin.github.io/reactable/index.html#demos) to see what is possible.
- A good example of a reactable that is also useful in its contents is [Nicola Rennie's Data Science Resources](https://nrennie.rbind.io/data-science-resources/).

### Movements example
The below is a relatively simple example that adds a search box, increases the default page side, changes some visual aspects of the table, and provides capitalized header values.

```{r}
#| label: movements-reactable
movements <- movements |> 
  arrange(date) |> 
  mutate(date = paste(day(date), # Transform to use of words for date
                      month(date, label = TRUE, abbr = FALSE),
                      year(date)))


reactable(movements,
          defaultPageSize = 15,
          searchable = TRUE,
          bordered = TRUE,
          highlight = TRUE,
          columns = list(
            person = colDef(name = "Person"),
            place = colDef(name = "Place"),
            date = colDef(name = "Date")
            )
          )
```

You can get quite creative with reactable, including adding [filtering widgets](https://glin.github.io/reactable/articles/examples.html#cross-widget-interactions) and a [button to download the data](https://glin.github.io/reactable/articles/examples.html#javascript-api) among many other options.